Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-06-25 paper/paper.tex
more revisions

  In this paper, we show their functional correspondence and how to syntactically 
  transform small-step abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, lightweight fusion, disentanglement, 
  refunctionalization, and finally to direct-style with delimited control. 
  Linearization expresses nondeterministic choices by first-order data types, 
  after which refunctionalization transforms the data type representing continuations
  into higher-order functions.
  Following the idea that in deterministic languages, reduction contexts of reduction
  semantics are defunctionalized continuations, we further apply this idea to nondeterministic
  AAM -- reduction contexts of it are refunctionalized to extended continuation-passing style.

@@

In the concrete world, the relations among reduction semantics, abstract machines,
definitional interpreters, and monadic interpreters have been intensively studied by
Danvy and his collaborators \cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202,
danvy2004refocusing, Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, 
Danvy:2006:RW:2171265.2171268, danvy2009towards, biernacka2009towards}.
The concrete abstract machines implement structural operational semantics in 
continuation-passing style, where the reduction contexts are defunctionalized continuations.
One can derive definitional interpreters by refunctionalizing the evaluation contexts of 
abstract machines, and by defunctionalizing the higher-order functions, one may obtain 
abstract machines in the reverse direction.

@@

The CESK machine is an abstract machine for describing semantics of and evaluating 
$\lambda$-calculus \cite{felleisen1987calculus}. The CESK machine models program execution 
as state transitions in a small-step fashion. As its name suggests, a machine state has 
four components:
1) \textit{Control} is the expression currently being evaluated.
2) \textit{Environment} is a map that contains the address of a variable in the lexical scope.
3) \textit{Store} models the heap of a program as a map from addresses to values.
  The address space consists of numbers (0-indexed).
  In our toy language, the only category of value is a closure, i.e., a function paired with
  an environment.
4) \textit{Continuation} represents the program stack. In this paper, we instantiate the 
  stack as a list of frames because the ANF simplifies the evaluation context. For direct 
  style $\lambda$-calculus, the continuations of CEK/CESK machines can be presented by
  variants of a data type.

@@

It is worth noting that the continuation class @Kont@ is defined as a list of
frames, where a frame can be considered an evaluation context in reduction semantics.
We represent frames using the @Frame@ class, which stores the information of a single
call-site, i.e., the information that can be used to resume the interrupted computation.
A @Frame@ constitutes a variable name @x@ to be bound later, and a control component
to which the program may resume, and its environment.

@@

As previously mentioned, the space of addresses must be finite in AAM.
Binding addresses are parameterized by variable names and the creation 
time of the binding, both of which are finite.
Continuation addresses @KAddr@ has two variants:
1) @Halt@ which corresponds to the empty stack, and
2) @ContAddr@ consists of the target expressions of callee, which are also finite.
%One may ask, why not also keep track of the time in continuation addresses?
%\todo{why target expression, allocation polyvariance}

@@

Given the space of addresses are finite, we can conclude that there are finite number
of environments and stores because the number of variables and closures are also finite.
This property guarantees a finite space of reachable states, and we can always have a
terminated analysis through Kleene's fixed-point iteration.

@@

In Danvy's paper \textit{Defunctionalized Interpreters for Programming
  Languages}, he mentions that for deterministic languages, a reduction semantics
is a structural operational semantics in continuation-passing style, where the reduction
context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
This poses a problem, as the underlying semantics of AAM is fundamentally nondeterministic.
But the evaluation context (i.e., the @Frame@ in our program) is \emph{not}
nondeterministic, and the worklist @todo@ actually implicitly handles the nondeterminism.
Thus, if we simply refunctionalize the frames to higher-order functions, 
it does not help us to move toward abstract definitional interpreters.

In order to address this problem, our first step of transformations
is to linearize all nondeterministic choices.
This step removes all nondeterminism from the @step@ function.
Likewise, the @Frame@ saves the information of the caller in concrete executions.
We then introduce another layer of control, and define a case class @NDCont@
that saves the information at a fork point when we have multiple target closures.
We also add a new field @ndk@ to the definition of state, which we now call @NDState@.
For clarity of presentation in differentiating between the two continuations, we elect to call
the first a \emph{normal continuation}, and the second a \emph{nondeterministic continuation}.

@@

Each @NDCont@ object contains a list of closures that are possible functions to be invoked,
a set of values that will be bound to the function's formal argument,
and the store, time, and frames at the fork point.
In the definition of @NDState@, an auxiliary function @toState@ is added that
converts itself to @State@.

@@

By atomically evaluating @f@, we obtain a set of closures, with the transition being 
deterministic in regards to the first element of the closure set.
As such, we prepare a new environment, a new store, and a new frame list only for the
first closure in that set.
A new nondeterministic continuation @new_ndk@ is also constructed,
which contains the rest of closures, the values of argument @ae@, the store,
the time, and the new frame list.
We use the store before updating, because for different closures they may form
different binding addresses @baddr@.
We also use the new frames, because all closures at this fork point share
the same stack and return point.

@@

However, since we have added nondeterminism into state, an empty list of frames means
we have reached the halt of \textit{one computation path}, and we should
determine the next state indicated by @NDCont@.
Thus, we have a pattern matching on @ndk@:
\begin{itemize}
  \item If the closure set is an empty set,
then we have tried all the closures of this fork point and should move to the
next fork point. Since the @NDCont@ is represented by a list, and we always
append new elements to its front, we are resuming to the most recent fork point by
popping up the front-most element.
  \item Otherwise, we pop a closure from the set, then build a new
environment and a new store for the body expression of the closure;
we use the same frame list and time that are copied from the fork point rather
than current one.
The nondeterministic continuation @new_ndk@ is also updated by the removal of that closure.
\end{itemize}

@@

With fusing complete, our AAM appears similar to a ``big-step'' interpreter, although it
still has the machine state representation inside. In the disentanglement transformation,
we identify the first-order data types which represent evaluation contexts
and lift the code blocks that handle these evaluation contexts to be individual functions.

Since there are two layers of continuations,
we obtain three mutually recursive functions: @drive_step@, which plays the same
role as before; @continue@, which is called from @drive_step@ when encountering an atomic 
expression, and handles the evaluation context (@Frame@) of the analyzed
language; and @ndcontinue@, which dispatches nondeterministic continuations @NDCont@,
and which is invoked from @continue@ when the normal stack is empty.

@@

Function @continue@ is mainly what we have seen for the second case in @drive_step@,
except that @ndcontinue@ is called when the normal continuation is empty.
Since we are analyzing a language in ANF, the first-order data representation of
continuations is simplified to a Scala @List@. An empty list @Nil@ represents the
halt of a computation path, whereas a cons @::@ means the top frame of the list
holds the binding variable and evaluation context.
An analogy to this in standard CEK machines would be an abstract class (e.g., @cont@)
with three variants: @halt@, @arg(value, cont)@, and @arg(exp, env, cont)@.

@@

For clarity, we first show how to refunctionalize normal continuations and
nondeterministic continuation without collecting semantics.
Then, adding an appropriate caching algorithm to make sure the analysis will
terminate at some fixed-point is a simple transformation of the program to cache-passing
style. Finally, we examine how computable pushdown control-flow analysis is established
through refunctionalization and caching.

@@

%TODO: better word to replace `accumulate`
Given a collection of elements of type @T@, an initial accumlated value of type 
@Ans@, a function @k@ that works on element's type @T@ and takes a continuation 
of type @Ans=>Ans@, and a function @m@ that works on and returns a value of type 
@Ans@ (which is also type @Ans=>Ans@), @nd@ chooses an element in the collection 
and applies @k@ to it, accumulates the value (@acc++ans@ where the @ans@ is the 
result comes from @k@) and eventually applies function @m@ on the accumulated 
value when there is no element in the collection.

@@

% TODO: explain more on how caches work
Finally, the @analyze@ function (as the entrance of the analysis) does a looping iteration
to find the fixed-point on caches starting from an empty cache.
If the @out@ cache of this iteration is equivalent to its @in@ cache, which means no more 
fact was discovered during this iteration, then we have reached a fixed-point that 
over-approximates the concrete evaluations, and thus can be returned.
Otherwise, the next iteration with the new @out@ cache installed at the @in@ cache position 
will be activated; in the meanwhile, an empty cache will be used as the @out@ cache.
Note that the continuation of the initial call to @aeval@ sets up the values for the
initial configuration in the final cache.

@@

In fact, what we started is an AAM with an unbounded stack which already precisely 
matches the calls and returns.
Refunctionalization makes the interpreter be a ``big-step'' semantic artifact, 
then the consequence is we have no places to save the context information. 
But refunctionalization not only forces us to sequentialize the order nondeterministic 
evaluation through higher-order functions representing continuations, 
but also drives us to use a new caching algorithm.

@@

Now we can use these two operators to redefine @aeval@. The abstract evaluator @aeval@ is
rewritten in direct-style, so that the last argument @cont@ disappears; we also add an 
annotation \verb|@cps[Ans]| on the return type to enable using delimited control operators 
in it.
Then, for every nondeterministic choice in the abstract interpreter, we can simply call 
@choices@ on the set of possible choices and write the program sequentially as concrete 
interpreters, and the subsequent statements and expressions become the delimited continuation
@f@. @choices@ also returns the latest cache to its left-hand side definition.
The cache is still accumulated to the next call to @choices@ or @aeval@, 
given the fact that the abstract interpreter should always use the latest cache.

@@

Over the years, Danvy and his collaborators applied refunctionalization and defunctionalization 
to many different abstract machines and interpreters and build functional correspondences among
them \cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing,
Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268,
danvy2009towards, biernacka2009towards}. Such semantic artifacts include the CEK machines,
the CLS machines, the SECD machines, call-by-need evaluators, monadic evaluators, etc.
Other applications are also found to be applicable, such as Dyck word recognizer, and 
Dijkstra's shunting-yard algorithm and even Quicksort algorithm \cite{Danvy:2006:RW:2171265.2171268}.

For deterministic languages, \citeauthor{Danvy:2008:DIP:1411204.1411206} shows a reduction semantics
is a structural operational semantics in continuation-passing style, where the reduction contexts
are defunctionalized (data type representing) continuations \cite{Danvy:2008:DIP:1411204.1411206}.
Not unexpectedly, the correspondence also exists for nondeterministic languages. For example,
\citeauthor{Danvy:2001:DW:773184.773202} show that refunctionalizing a regualr expressions matcher yields
its counterpart with two layer of continuations \cite{Danvy:2001:DW:773184.773202}. And the multi-layer
continuation-passing style is the underlying semantics of control operatos like @reset@ and @shift@
\cite{Danvy:1990:AC:91556.91622}.

Our work applies these ideas to the nondeterministic AAM. The additional layer of continuation controls
the nondeterministic choice over closures. After identifying and explicitly exposing (Section~\ref{linear})
that layer of nondeterministic continuations, the whole program can be refunctionalized to extended
continuation-passing style, and then back to direct-style by the use of delimited control operators.

>>>>

  In this paper, we show their functional correspondence and how to syntactically 
  transform small-step abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, lightweight fusion, disentanglement, 
  refunctionalization, and finally back to direct-style with delimited control operators. 
  Linearization expresses nondeterministic choices by first-order data types, 
  after which refunctionalization transforms the data type representing continuations
  into higher-order functions.
  Following the idea that in deterministic languages, reduction contexts of reduction
  semantics are defunctionalized continuations, we further apply this idea to nondeterministic
  AAM -- its reduction contexts are refunctionalized to extended continuation-passing style.

@@

In the concrete world, the relations among reduction semantics, abstract machines,
definitional interpreters, and monadic interpreters have been intensively studied by
Danvy and his collaborators \cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202,
danvy2004refocusing, Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, 
Danvy:2006:RW:2171265.2171268, danvy2009towards, biernacka2009towards}.
The concrete abstract machines implement structural operational semantics in 
continuation-passing style, where the reduction contexts are defunctionalized continuations.
One can derive definitional interpreters by refunctionalizing the reduction contexts of 
abstract machines, and by defunctionalizing the higher-order functions, one may obtain 
abstract machines in the reverse direction.

@@

The CESK machine is an abstract machine for describing semantics of and evaluating 
$\lambda$-calculus \cite{felleisen1987calculus}. The CESK machine models program execution 
as state transitions in a small-step fashion. As its name suggests, a machine state has 
four components:
1) \textit{Control} is the expression currently being evaluated.
2) \textit{Environment} is a map that contains the address of a variable in the lexical scope.
3) \textit{Store} models the heap of a program as a map from addresses to values.
  The address space consists of numbers (0-indexed).
  In our toy language, the only category of value is a closure, i.e., a function paired with
  an environment.
4) \textit{Continuation} represents the program's call stack. In this paper, we instantiate the 
  stack as a list of frames because the ANF simplifies the evaluation context. For direct % TODO
  style $\lambda$-calculus, the continuations of CEK/CESK machines can be presented by
  variants of a data type.

@@

It is worth noting that the continuation class @Kont@ is defined as a list of
frames, where a frame can be considered a reduction context in reduction semantics.
We represent frames using the @Frame@ class, which stores the information of a single
call-site, i.e., the information that can be used to resume the interrupted computation.
A @Frame@ constitutes a variable name @x@ to be bound later, and a control component
to which the program may resume, and its environment.

@@

As previously mentioned, the space of states is finite when the space of addresses is finite.
To make this happen, binding addresses are parameterized by variable names and the creation
time of the binding, both of which are finite. Continuation addresses @KAddr@ has two variants:
1) @Halt@ which corresponds to the empty stack, and
2) @ContAddr@ consists of the target expressions of callee, which are also finite.
%One may ask, why not also keep track of the time in continuation addresses?
%\todo{why target expression, allocation polyvariance}

@@

Given that the space of addresses is finite, we can conclude that there are finite number
of environments and stores because the numbers of variables and closures are also finite.
This property guarantees a finite space of reachable states, and we can always have a
terminated analysis through Kleene's fixed-point iteration.

@@

The underlying semantics of AAM is fundamentally nondeterministic: there are possibly
multiple target closures when dereferencing an address, and we have to explore both branches
if the conditionals exist in the language. But the reduction context (i.e., the @Frame@ in
our program) is \emph{not} nondeterministic, and the worklist @todo@ actually implicitly
handles the nondeterminism. Thus, if we simply refunctionalize the frames to higher-order
functions, it does not help us to move toward abstract definitional interpreters.

In Danvy's paper \textit{Defunctionalized Interpreters for Programming Languages},
he mentions that for deterministic languages, a reduction semantics is a structural
operational semantics in continuation-passing style, where the reduction context is
a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}. Therefore
refunctionalizing a deterministic language yields an interpreter in continuation-passing
style; nevertheless, doing so for nondeterministic languages yields an interpreter
with multiple layers of continuation \cite{Danvy:2006:RW:2171265.2171268}.

We apply this idea to AAM and our first step of transformations is to linearize all
nondeterministic choices. This step removes all nondeterminism from the @step@ function.
Likewise, the @Frame@ saves the information of the caller in concrete executions.
We then introduce another layer of control, and define a case class @NDCont@
that saves the information at a fork point when we have multiple target closures.
We also add a new field @ndk@ to the definition of state, which we now call @NDState@.
For clarity of presentation in differentiating between the two continuations, we elect to call
the first a \emph{normal continuation}, and the second a \emph{meta-continuation}
\footnote{Traditionally, they are also called \emph{success continuation} and \emph{failure
continuation} \cite{Danvy:1990:AC:91556.91622}.}.

@@

Here we also use first-order data types to explicitly represent the meta-continuation which
controls the nondeterminism.
Each @NDCont@ object contains a list of closures that are possible functions to be invoked,
a set of values that will be bound to the function's formal argument, and the store, time,
and frames at the fork point. In the definition of @NDState@, an auxiliary function @toState@
is added that converts itself to @State@.

@@

By atomically evaluating @f@, we obtain a set of closures, with the transition being 
deterministic in regards to the first element of the closure set.
As such, we prepare a new environment, a new store, and a new frame list only for the
first closure in that set. A new meta-continuation @new_ndk@ is also constructed,
which contains the rest of closures, the values of argument @ae@, the store, the time,
and the new frame list. We use the store before updating, because for different closures
they may form different binding addresses @baddr@. We also use the new frames, because
all closures at this fork point share the same stack and return point.

@@

However, since we have added nondeterminism into state, an empty list of frames means
we have reached the halt of \textit{one computation path}, and we should determine the
next state indicated by @NDCont@. Thus, we have a pattern matching on @ndk@:

\begin{itemize}
  \item If the closure set is an empty set,
then we have tried all the closures of this fork point and should move to the
next fork point. Since the @NDCont@ is represented by a list, and we always
append new elements to its front, we are resuming to the most recent fork point by
popping up the front-most element.
  \item Otherwise, we pop a closure from the set, then build a new
environment and a new store for the body expression of the closure;
we use the same frame list and time that are copied from the fork point rather
than current one.
The meta-continuation @new_ndk@ is also updated by the removal of that closure.
\end{itemize}

@@

With fusing complete, our AAM appears similar to a ``big-step'' interpreter, although it
still has the machine state representation inside. In the disentanglement transformation,
we identify the first-order data types which represent reduction contexts
and lift the code blocks that handle these reduction contexts to be individual functions.

Since there are two layers of continuations, we obtain three mutually recursive functions:
@drive_step@, which plays the same role as before; @continue@, which is called from
@drive_step@ when encountering an atomic expression, and handles the reduction context
(@Frame@) of the analyzed language; and @ndcontinue@, which dispatches meta-continuations
@NDCont@, and which is invoked from @continue@ when the normal stack is empty.

@@

Function @continue@ is mainly what we have seen for the second case in @drive_step@,
except that @ndcontinue@ is called when the normal continuation is empty.
Since we are analyzing a language in ANF, the first-order data representation of
continuations is simplified to a Scala @List@. An empty list @Nil@ represents the %TODO
halt of a computation path, whereas a cons @::@ means the top frame of the list
holds the binding variable and context.
An analogy to this in standard CEK machines would be an abstract class (e.g., @cont@)
with three variants: @halt@, @arg(value, cont)@, and @arg(exp, env, cont)@.

@@

For clarity, we first show how to refunctionalize normal continuations and
meta-continuation without collecting semantics.
Then, adding an appropriate caching algorithm to make sure the analysis will
terminate at some fixed-point is a simple transformation of the program to cache-passing
style. Finally, we examine how computable pushdown control-flow analysis is established
through refunctionalization and caching.

@@

Given a collection of elements of type @T@, an initial accumlator of type 
@Ans@, a function @k@ that works on element's type @T@ and takes a continuation 
of type @Ans=>Ans@, and a function @m@ that works on and returns a value of type 
@Ans@ (which is also type @Ans=>Ans@), @nd@ chooses an element in the collection 
and applies @k@ to it, accumulates the value (@acc++ans@ where the @ans@ is the 
result comes from @k@) and eventually applies function @m@ on the accumulated 
value when there is no element in the collection.

@@

% TODO: explain more on how caches work
Finally, the @analyze@ function (as the entrance of the analysis) does a looping iteration
to find the fixed-point on caches starting from an empty cache.
If the @out@ cache of this iteration is equivalent to its @in@ cache, which means no more 
fact was discovered during this iteration, then we have reached a fixed-point that 
over-approximates the concrete evaluations, and thus the result can be returned.
Otherwise, the next iteration with the new @out@ cache installed at the @in@ cache
position will be activated; in the meanwhile, an empty cache will be used as the @out@
cache. Note that the continuation of the initial call to @aeval@ sets up the values for
the initial configuration in the final cache.

@@

In fact, what we started is an AAM with an unbounded stack which already precisely 
matches the calls and returns.
Refunctionalization makes the interpreter be a ``big-step'' semantic artifact, 
then the consequence is we have no places to save the context information. 
But refunctionalization not only forces us to sequentialize the order of nondeterministic 
evaluation through higher-order functions representing continuations, but also drives us
to use a new caching algorithm.

@@

Now we can use these two operators to redefine @aeval@. The abstract evaluator @aeval@ is
rewritten in direct-style, so that the last argument @cont@ disappears; we also add an 
annotation \verb|@cps[Ans]| on the return type to enable using delimited control operators 
in it.
Then, for every nondeterministic choice in the abstract interpreter, we can simply call 
@choices@ on the set of possible choices and write the program sequentially as concrete 
interpreters, and the subsequent statements and expressions become the delimited continuation
@f@. The function @choices@ also returns the latest cache to its left-hand side when invoking.
The cache will be still accumulated to the next call to @choices@ or @aeval@, 
given the fact that the abstract interpreter should always use the latest cache.

@@

Over the years, Danvy and his collaborators applied refunctionalization and defunctionalization 
to many different abstract machines and interpreters and build functional correspondences among
them \cite{Ager:2003:FCE:888251.888254, Danvy:2001:DW:773184.773202, danvy2004refocusing,
Danvy:2008:DIP:1411204.1411206, AGER2004223, ager2005functional, Danvy:2006:RW:2171265.2171268,
danvy2009towards, biernacka2009towards}. Such semantic artifacts include the CEK machines,
the CLS machines, the SECD machines, call-by-need evaluators, monadic evaluators, etc.
Other applications are also found to be applicable, such as Dyck word recognizer, and 
Dijkstra's shunting-yard algorithm and even Quicksort algorithm \cite{Danvy:2006:RW:2171265.2171268}.
For deterministic languages, \citeauthor{Danvy:2008:DIP:1411204.1411206} shows a reduction semantics
is a structural operational semantics in continuation-passing style, where the reduction contexts
are defunctionalized (data type representing) continuations \cite{Danvy:2008:DIP:1411204.1411206}.
Not unexpectedly, the correspondence also exists for nondeterministic languages. For example,
\citeauthor{Danvy:2001:DW:773184.773202} show that refunctionalizing a regualr expressions matcher yields
its counterpart with two layer of continuations \cite{Danvy:2001:DW:773184.773202}.
And the multi-layer continuation-passing style is the underlying semantics of control operatos
like @reset@ and @shift@ \cite{Danvy:1990:AC:91556.91622}.

Our work applies these ideas to the nondeterministic AAM. The additional layer of continuation controls
the nondeterministic choice over closures. After identifying and explicitly exposing (Section~\ref{linear})
that layer of meta-continuations, the whole program can be refunctionalized to extended
continuation-passing style, and then back to direct-style by the use of delimited control operators.
